Distributional Trees for Circular Data

### 2019-02-01 pkg/disttree/inst/IWSM2019/abstract_IWSM2019.tex
Adapt application
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2421 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 479ef4939d3a38033574fef93486317a6a839943

%\begin{figure}
%\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
%\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
%\label{fig:tree}
%\end{figure}

@@

As an illustration for the tree based forecast performance of predicting wind
directions at Innsbruck airport, a BIC based post-pruned regression tree of
depth three is shown in Fig.~\ref{fig:tree}. According to the fitted von Mises
distributions three distinct wind regimes: Firstly, up-valley winds blowing
from the valley mouth towards the upper valley (from east to west) as shown in
the final nodes 4 and 5. Secondly, downslope winds blowing across the Alpine
crest along the Wipp Valley towards Innsbruck (from south-east to north-west)
as predicted in nodes 7 and 8. Thirdly, down-valley winds blowing in the
direction of the valley mouth (from west to east) present in nodes 12, 14, and
15. The final node 11 with a density parameter $\kappa$ of $0.84$ captures the
cases with rather low wind speeds and therefore no clear wind direction.
Generally, the splits in the lagged wind direction perform a clustering into
the different wind regimes. As winds typically blow from areas of high pressure
to areas of low pressure and have larger wind speeds in case of stronger
pressure gradients, these respective splits in pressure gradients and wind
speed divide the predictions according to the stability of the lagged wind
direction. As an example node 3 differentiates in the pressure difference along
the valley, and the cases with a stronger pressure gradient have a higher
fitted $\kappa$ as indicator for the stability of the forecast. In other words,
in final nodes with a higher $\kappa$ most training cases are classified
correctly and therefore the density parameter is an indicator for the
uncertainty of the forecast.

>>>>

\begin{figure}
\includegraphics[height = .5\textheight,angle=90,origin=c]{circtree_plot_big.pdf}
\caption{Fitted distribution tree for a circular response employing the von Mises distribution.}
\label{fig:tree}
\end{figure}

@@

As an illustration of the tree based wind direction forecasts at Innsbruck airport, a
BIC based post-pruned regression tree of depth three is shown in
Fig.~\ref{fig:tree}. Three distinct wind regimes are fitted within the final
nodes: Firstly, up-valley winds blowing from the valley mouth towards the upper
valley (from east to west, as predicted in the final nodes 4 and 5). Secondly,
downslope winds blowing across the Alpine crest along the Wipp Valley towards
Innsbruck (from south-east to north-west, as predicted in nodes 7 and 8).
Thirdly, down-valley winds blowing in the direction of the valley mouth (from
west to east, as predicted in nodes 12, 14, and 15). The final node 11 with a
concentration parameter $\kappa$ of only $0.84$ captures the cases with rather low
wind speeds and typically no clear wind direction or distinct wind regime.
In general, winds typically blow from areas of high pressure to areas of low
pressure, and a strong pressure gradient leads to large wind speeds and a stable
wind direction. Therefore, the splits in the lagged wind direction perform a
clustering into the different wind regimes and the further splits in pressure
gradients or wind speed divide the predictions according to the consistency of
the lagged wind direction. This can be seen in the final nodes with strong
pressure gradients or high wind speeds where most training cases are classified
correctly and therefore the concentration parameter as an indicator of the
certainty of the forecast is rather large.
